Refining from the BON High-level Specification Language to the VHDL-AMS Hardware Description Language

### 2009-09-04 papers/BON_VHDL/paper.tex
reorganized paper

% \usepackage{html}
% \usepackage{url}
\usepackage{xspace}
%\usepackage{doublespace}
\usepackage{tabularx}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{eucal}
\usepackage{stmaryrd}
% \ifpdf
% \usepackage[centredisplay]{diagrams}
% \else
% \usepackage[centredisplay,PostScript=dvips]{diagrams}
% \fi
\usepackage{float}

\ifpdf
\usepackage[pdftex,bookmarks=false,a4paper=false,
            plainpages=false,naturalnames=true,
            colorlinks=true,pdfstartview=FitV,
            linkcolor=blue,citecolor=blue,urlcolor=blue,
            pdfauthor="Joseph R. Kiniry and Dragan Stosic"]{hyperref}
\else
\usepackage[dvips,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\fi
\ifpdf
  \DeclareGraphicsRule{*}{mps}{*}{}
\fi

@@

\lstdefinelanguage{VHDL} {
  morekeywords={not,xnor,and,nand, nor,or,xor,integer,byte,real},
  morekeywords={character,enum,type,boolean,bit,time,null,std_ulogic},
  morekeywords={sll, srl, sla, sra, rol, ror},
  morekeywords={rem,abs,mod},
  morekeywords={terminal,T'reference, T'contribution,T'tolerance},
  morekeywords={reference,contribution},
  morekeywords={nature,subnature,electrical,magnetic,translational,translational_v},
  morekeywords={rotational,rotational_v,fluidic,thermal,radiant},
  morekeywords={voltage,angular_velocity,pressure,velocity},
  morekeywords={current,force,torque,volumetric,flow_rate},
  morekeywords={T'left,T'right,T'high,T'ascending,T'image(x),T'value(s)},
  morekeywords={left,right,high,ascending,image,value},
  morekeywords={T'pos(s),T'val(x),T'succ(x),T'pred(x)},
  morekeywords={pos,val,succ,pred},
  morekeywords={A'left(n),A'right(n),A'low(n),A'high(n)},
  morekeywords={left,right,low,high},
  morekeywords={A'range(n),A'reverse_range(n),A'length(n),A'ascending(n)},
  morekeywords={range,reverse_range,length,ascending},
  morekeywords={T'leftof(x),T'rightof(x),T'base},
  morekeywords={leftof,rightof,base},
  morekeywords={Q'tolerance, Q'above,Q'delayed,Q'dot,Q'integ},
  morekeywords={delayed,dot,integ},
  morekeywords={Q'slew,Q'ztf,Q'ltf,Q'zoh},
  morekeywords={N'across, N'through},
  morekeywords={across,through,zoh,ltf,slew,ztf},
  morekeywords={above,tolerance,group,quantity},
  morecomment=[l]{--},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
 literate=,
  showstringspaces=false
}

@@

The \STs, as an embedded system, have several unusual built-in sensors
and emitters, sensor nodes can be extended via a general-purpose USB bus,
and \STs contain an FPGA that is dynamically updated on a per-experiment
basis.  \STs are custom-designed in VHDL-AMS.  Like many hardware-based systems,
 to write new software for or against the \ST architecture one either uses
the actual hardware or one runs a \STS.  But this simulator emulates a
piece of hardware, one would like to know if the simulator
\emph{actually} simulates the hardware.  We have developed an approach
through which we refine the VHDL specification up to the formal BON
specification language, and from this specification we generate an
executable, functional model-based specification in the JML language.
We have used this specification to implement a fully validated and
verified \ST simulator.

@@

%======================================================================
\section{Introduction}
\label{sec:introduction}
The \ST sensor platform, represents heterogeneous collection of digital, analog
and mixed signal hardware circuits. Embedded system complexity
requires the need to strive for formal descriptions each component
separately in order to produce specification of the \STS.
Our aim is to be able to precisely describe the structures of circuits,
the behavior of circuits and the formal link between structures and behaviors.
In other words, the behavioral specification of the \STS may be
represented as a \inv on the signals at the circuit's ports and may be obtained
by taking the conjunction of the \emph{invariants} that describe
behaviors of the individual components separately.
For this purpose we need  sophisticated tool and methodology
in order to be able to achieve this goal. 
The authors will illustrate in this article how BON can represent a high
level analogue mapping and will further illustrate a procedure for
transformation from VHDL-AMS to BON.\\
This article provides three main contributions.
\begin{enumerate}
\item
 It provides a solution how to formally emulate
 piece of hardware using BON-JML approach; up to date BON and JML
 have been used mostly in the pure object oriented software construction.   
\item
It relates hardware description code and formal calculus.
This means that if there  is a code that is "correct by construction",
then it will be derivable from BON (informal/formal) specification,
via the refinement calculus to JML implementation.
\end{enumerate}
The rest of this article is organized as follows.
In \autoref{sec:vhdlams} we have  presented an overview of VHDL-AMS language.
\autoref{sec:choiceobon} defines  BON principles and explains the
reasons why we have decided to use BON method.
In \autoref{sec:mappingVB} we will show mapping between VHDL-AMS and BON.
A comprehensive list of the mapping between BON and VHDL-AMS can be
found in the appendix. 
The central part of the our article is \autoref{sec:refVB}, in which
we will give practical aspect of refinement, applied on lowpass filter, 
in order to produce formal specification and software simulation.\\ 
%=====================================================================

@@

\begin{itemize}
\item
\emph{Terminals} represents phisical connection points or circuit nodes
of a network connectivity. They are declared to be of various
\emph{natures}, which represent different energy domains of a system.
Terminals hold no values; their only role is to facilitate the formation
of conservative equation sets that, in turn, constrain the associated
\emph{branch quatities}.
\item
The \emph{quantities} denote a waveform or a time series of values.
The behaviour of the \emph{quantities} is given by the \emph{simultaneous
 statements}. They take values as a result of solving the set of simultaneous
ordinary differential and algebraic equations (DAEs). To simplify,
quantities can be understood as analog unknowns of (DAEs) that specify
analog behaviour. There are three kinds of quantity in VHDL-AMS: free,
branch and source quantities. A \emph{free quantity} can be used in
signal-flow modeling. A \emph{branch quantity} is used to model conservative energy
systems. A \emph{source quantity} is used for frequency and noise modeling.
\item
The \emph{simultaneous statements} creates one or more characteristic
equations that are solved by the dedicated simulation kernel a.k.a.
"analog solver", to model continuously valued behaviour. 
\item
The \emph{procedural statements} or \emph{procedural} express analog behaviour
using sequential statements. Actually procedural describes analog behaviour
by computing new values for quantities from the current value of quantities.
\item
The \emph{break statement} is an construct which allows to
initialize the DAE system at the beginning of a transient
simulation and to mark discontinuities during such a transient
simulation.For any concurrent break statement there is an associated process.
A break statement allows restart of the analog kernel,
than overrides quantity values and executes concurrently with the analogue model.

@@

a "real capacitor" at the level of individual components (a resistor and
ideal capacitor) and their behaviours, or at the more abstract level through

@@

The main theme of this article is the mapping between VHDL-AMS and BON.
\section{Refinement from VHDL-AMS to BON }
\label{sec:refVB}
\includegraphics{circ.1}

%=====================================================================
\section{Conclusion}

@@

%======================================================================
\appendix
\newpage
\section {APPENDIX}
\input{static}
\input{arith}
\input{shift}
\input{relation}
\input{logic}
\input{types}
\input{assignment}
\input{iscaltypeattr}
\input{tolerance}
\input{effort}
\input{flow}
\input{nature}
\input{iscalnature}
\input{iarrayattr}
\input{scalnat}
\input{iterminal}
\input{terminal}
\input{iquantity}
\input{quantity}
\input{quantityport}
\input{terminalport}
\input{signal}
\input{signalattr}
\input{wait}
%======================================================================

>>>>

% \usepackage{html}
% \usepackage{url}
\usepackage{xspace}
%\usepackage{doublespace}
\usepackage{tabularx}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{eucal}
\usepackage{stmaryrd}
% \ifpdf
% \usepackage[centredisplay]{diagrams}
% \else
% \usepackage[centredisplay,PostScript=dvips]{diagrams}
% \fi
\usepackage{float}
\usepackage{subfig}

\ifpdf
\usepackage[pdftex,bookmarks=false,a4paper=false,
            plainpages=false,naturalnames=true,
            colorlinks=true,pdfstartview=FitV,
            linkcolor=blue,citecolor=blue,urlcolor=blue,
            pdfauthor="Joseph R. Kiniry and Dragan Stosic"]{hyperref}
\else
\usepackage[dvips,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\fi
\ifpdf
  \DeclareGraphicsRule{*}{mps}{*}{}
\fi
\def\figureautorefname{Fig.}

@@

\lstdefinelanguage{VHDL} {
  morekeywords={not,xnor,and,nand, nor,or,xor,integer,byte,real},
  morekeywords={character,enum,type,boolean,bit,time,null,std_ulogic},
  morekeywords={sll, srl, sla, sra, rol, ror},
  morekeywords={rem,abs,mod},
  morekeywords={terminal,T'reference, T'contribution,T'tolerance},
  morekeywords={reference,contribution},
  morekeywords={nature,subnature,electrical,magnetic,translational,translational_v},
  morekeywords={rotational,rotational_v,fluidic,thermal,radiant},
  morekeywords={voltage,angular_velocity,pressure,velocity},
  morekeywords={current,force,torque,volumetric,flow_rate},
  morekeywords={T'left,T'right,T'high,T'ascending,T'image(x),T'value(s)},
  morekeywords={left,right,high,ascending,image,value},
  morekeywords={T'pos(s),T'val(x),T'succ(x),T'pred(x)},
  morekeywords={pos,val,succ,pred},
  morekeywords={A'left(n),A'right(n),A'low(n),A'high(n)},
  morekeywords={left,right,low,high},
  morekeywords={A'range(n),A'reverse_range(n),A'length(n),A'ascending(n)},
  morekeywords={range,reverse_range,length,ascending},
  morekeywords={T'leftof(x),T'rightof(x),T'base},
  morekeywords={leftof,rightof,base},
  morekeywords={Q'tolerance, Q'above,Q'delayed,Q'dot,Q'integ},
  morekeywords={delayed,dot,integ},
  morekeywords={Q'slew,Q'ztf,Q'ltf,Q'zoh},
  morekeywords={N'across, N'through},
  morekeywords={across,through,zoh,ltf,slew,ztf},
  morekeywords={above,tolerance,group,quantity},
  morekeywords={entity,end,use,port,generic,is,library,all},
  morekeywords={architecture,of,through,begin,constant},
  morecomment=[l]{--},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
 literate=,
  showstringspaces=false
}

@@

The \STs, as an embedded system, have several unusual built-in sensors
and emitters, sensor nodes can be extended via a general-purpose USB bus,
and \STs contain an FPGA that is dynamically updated on a per-experiment
basis.  \STs are custom-designed in VHDL-AMS.  Like many hardware-based systems,
 to write new software for or against the \ST architecture one either uses
the actual hardware or one runs a \STS.  But this simulator emulates a
piece of hardware, one would like to know if the simulator
\emph{actually} simulates the hardware.  We have developed an approach
through which we refine the VHDL-AMS specification up to the formal BON
specification language, and from this specification we generate an
executable, functional model-based specification in the JML language.
We have used this specification to implement a fully validated and
verified \ST simulator.

@@

%======================================================================
\section{Introduction}
\label{sec:introduction}
The \ST sensor platform, represents heterogeneous collection of digital, analog
and mixed signal hardware circuits. Embedded system complexity
requires the need to strive for formal descriptions each component
separately in order to produce specification of the \STS.
Our aim is to be able to describe the structures of circuits,
the behavior of circuits and the formal link between structures and behaviors.
In other words, the behavioral specification of the \STS may be
represented as a \inv on the signals at the circuit's ports and may be obtained
by taking the conjunction of the \emph{invariants} that describe
behaviors of the individual components separately.
For this purpose we need  sophisticated tool and methodology
in order to be able to achieve this goal. 
The authors will illustrate in this article how BON can represent a high
level analogue mapping and will further illustrate a procedure for
transformation from VHDL-AMS to BON.\\
This article provides two main contributions.
\begin{enumerate}
\item
 It provides a solution how to formally emulate
 piece of hardware using BON-JML approach; up to date BON and JML
 have been used mostly in the pure object oriented software construction.   
\item
It relates hardware description code and refinement.
This means that if there  is a code that is "correct by construction",
then it will be derivable from BON (informal/formal) specification to 
JML implementation.
\end{enumerate}
The rest of this article is organized as follows.
In \autoref{sec:vhdlams} we have  presented an overview of VHDL-AMS language.
\autoref{sec:choiceobon} defines  BON principles and explains the
reasons why we have decided to use BON method.
In \autoref{sec:mappingVB} we will show mapping between VHDL-AMS and BON.
A comprehensive list of the mapping between BON and VHDL-AMS can be
found in the appendix. 
The central part of the our article is \autoref{sec:refVB}, in which
we will give practical aspect of refinement , applied on 
first order passive RC Low-Pass filter, in order to produce formal 
specification and software simulation.\\ 
%=====================================================================

@@

\begin{itemize}
\item
\emph{Terminals} represents phisical connection points or circuit nodes
of a network connectivity. They are declared to be of various
\emph{natures}, which represent different energy domains of a system.
Terminals hold no values; their only role is to facilitate the formation
of conservative equation sets that, in turn, constrain the associated
\emph{branch quatities}.
\item
The \emph{quantities} denote a waveform or a time series of values.
The behaviour of the \emph{quantities} is given by the \emph{simultaneous
 statements}. They take values as a result of solving the set of simultaneous
ordinary differential and algebraic equations (DAEs). To simplify,
quantities can be understood as analog unknowns of (DAEs) that specify
analog behaviour. There are three kinds of quantity in VHDL-AMS: free,
branch and source quantities. A \emph{free quantity} can be used in
signal-flow modeling. A \emph{branch quantity} is used to model conservative energy
systems. A \emph{source quantity} is used for frequency and noise modeling.
\item
The \emph{simultaneous statements} creates one or more characteristic
equations that are solved by the dedicated simulation kernel a.k.a.
"analog solver", to model continuously valued behaviour. 
\item
The \emph{procedural statements} or \emph{procedural} express analog behaviour
using sequential statements. Actually procedural describes analog behaviour
by computing new values for quantities from the current value of quantities.
\item
The \emph{break statement} is an construct which allows to
initialize the DAE system at the beginning of a transient
simulation and to mark discontinuities during such a transient
simulation. For any concurrent break statement there is an associated process.
A break statement allows restart of the analog kernel,
than overrides quantity values and executes concurrently with the analogue model.

@@

a "lowpass-filter" at the level of individual components (a resistor and 
capacitor) and their behaviours, or at the more abstract level through

@@

\section{Refinement from VHDL-AMS to BON }
\label{sec:refVB}
The main theme of this article is the translation from VHDL-AMS to BON using 
refinement approach.
We will refine the VHDL-AMS code up to the formal BON, and from this 
specification we will generate an executable specification in the JML language.
For this purpose, we will demonstrate our approach in BON-JML fashion trought 
low-pass filter example.
\input{lowpassRC}
We can implement the last equation (see \autoref{fig:lowpassfig}) using the \texttt{'dot} 
implicite attribute of quantity. For this purpose we introduce relation 
\texttt{$s \rightarrow d/dt \rightarrow 'dot$} . The transformation looks as follows: \\
\begin{equation}
v_{in} = v_{out} + \tau_psv_{out} 
   \begin{array}[b]{c}                                                                                                                                    
      {\scriptscriptstyle s \rightarrow d/dt}  \\=                                                                                                                  
   \end{array} 
v_{out} + \tau_p\cfrac{dv_{out}}{dt}
\begin{array}[b]{c}                                                                                                                                    
      {\scriptscriptstyle d/dt \rightarrow } {\scriptscriptstyle 'dot} \\=                                                                                                                  
   \end{array}
  v_{in} = v_{out} + \tau_pv_{out}{'dot} 
\end{equation} 
\input{lowpass}
We have not yet said what specification and refinement are.
Let us introduce these therms first.\\ 
A specification of a program component is a contract that gives some constraints 
on how the component is to behave but leaves freedom for the implementator to 
decide how the actual behavior of the component is to be realized.
Specification involves \emph{choice} between a number of alternatives; this 
relationship is one-to-many. As we have shown in \autoref{tab:rescap}, corresponding 
informal specifications represents high level of abstraction derived from VHDL-AMS code.

A refinement represents the concept that can be understud as series of 
correctness-preserving transformations from general perspective to 
a precise level of detail. Accurately, if we compare two specifications,
\texttt{S} and \texttt{S'}, then we can conclude that the \texttt{S} 
is more abstract than \texttt{S'} if any condition that we can establish 
with the first specification can also be established by second. 
It means that "\texttt{S is \emph{refined} by S'}", 
and we denote this  as \texttt{S $\rightsquigarrow$ S'}.
Let say that the first informal BON specification is denoted by 
\texttt{S} (see \autoref{tab:rescap}), and second, which is given 
in Table2  is denoted by \texttt{S'}.
In this case, the refinement relation represents an \emph{isomorphism} 
and maps the following:
%=====================================================================
\section{Conclusion}

@@

%======================================================================
%\appendix
%\newpage
%\section {APPENDIX}
%\input{static}
%\input{arith}
%\input{shift}
%\input{relation}
%\input{logic}
%\input{types}
%\input{assignment}
%\input{iscaltypeattr}
%\input{tolerance}
%\input{effort}
%\input{flow}
%\input{nature}
%\input{iscalnature}
%\input{iarrayattr}
%\input{scalnat}
%\input{iterminal}
%\input{terminal}
%\input{iquantity}
%\input{quantity}
%\input{quantityport}
%\input{terminalport}
%\input{signal}
%\input{signalattr}
%\input{wait}

%======================================================================
